---
title: "Coding Assignment 2"
output: pdf_document
---

*Objectives of this assignment:*

-   Gain familiarity in working in R notebooks.

-   Review some beginning information about data types and data objects.

-   Gain introduction to errors and use knowledge to diagnose problem.

\ 
*Some notes about running chunks one by one:*
-   To run a single chunk, you can click on the green triangle pointing right within the chunk. 
-   To run all chunks above a given chunk, you press the gray triangle pointing down with the green dash underneath. 
-   Notice that at the top of the pane, in the right corner, there is also a drop down menu for different run options. It also shows keyboard shortcuts. Play around with these different options.
-   When you have tested and are finished with all text and all code chunks, click "Knit" at the top of the pane. This will run all chunks and "compile" a PDF that includes text, code chunks and output. 

\

The following code chunk assigns a vector to the object `run.times1`. Note that `run.times1` holds data on how long a person ran each of seven days in a week.

```{r hw2chunk1}
run.times1 <- c(47.25,46.50,50.25,44.75,45.00,31.50,48.50)
```

What is the "data type" of `run.times1`?

[TYPE YOUR ANSWER HERE, 1 PT]

What is the "data structure" of `run.times1`?

[TYPE YOUR ANSWER HERE, 1 PT]

In the following code chunk, write a line of code that will display the information in the object `run.times1`?

```{r hw2chunk2}
run.times1
```

```{r hw2chunk3}
run.times2 <- c(46.50,50.25,0,30.75,27.00,48.25,56.00)
run.times.by.week <- cbind(run.times1,run.times2)
run.times.by.week
```
