---
title: "Coding Assignment 2"
output: pdf_document
---

### Objectives of this assignment

-   Gain familiarity in working in R notebooks.

-   Review introductory material about R data types and R data objects.

-   Gain introduction to errors and use knowledge from class to diagnose problem.

 

### Some notes about running code chunks

-   To run a single chunk, you can click on the green triangle pointing right within the chunk.

-   To run all chunks above a given chunk, you press the gray triangle pointing down with the green dash underneath.

-   Notice that at the top of the pane, in the right corner, there is also a drop down menu for different run options. It also shows keyboard shortcuts. Play around with these different options.

-   When you have tested and are finished with all text and all code chunks, click "Knit" at the top of the pane. This will run all chunks and "compile" a PDF that includes text, code chunks and output.

### 

### Coding

1.  The following code chunk assigns a vector to the object `run.times1`. Note that `run.times1` holds data on how long a person ran each of seven days in a week.

```{r hw2chunk1}
run.times1 <- c(47.25,46.50,50.25,44.75,45.00,31.50,48.50)
```

-   What is the "data type" of `run.times1`?

[TYPE YOUR ANSWER HERE, 1 POINT]

-   What is the "data structure" of `run.times1`?

[TYPE YOUR ANSWER HERE, 1 POINT]

-   Below, create a new code chunk. Name the code chunk `hw2chunk2`. In it, write a line of code that will display (print) the the object `run.times1`?

[CREATE A CODE CHUNK HERE, 2 POINTS]

2.  The object `run.times.by.week` shows data collected on someone's run times over two weeks.

```{r hw2chunk3}
run.times2 <- c(46.50,50.25,0,30.75,27.00,48.25,56.00)
run.times.by.week <- cbind(run.times1,run.times2)
run.times.by.week
```

-    What are two observations that tell a story about the data?

[ENTER YOUR ANSWER HERE, 2 POINTS FOR COMPLETION, NO WRONG ANSWERS]

-   What are two analyses you do with this run time data when you know how to do it in R?

[ENTER YOUR ANSWER HERE, 2 POINTS FOR COMPLETION, NO WRONG ANSWERS]

-   In the following code chunk, write a line of code to show the number of rows and columns in the matrix `run.times.by.week`. Also, add a comment to the code chunk, saying whatever you think makes sense. [2 POINTS]

```{r hw2chunk4}

```

3.  Note that when we use double equal signs (`==`) in R, we are evaluating whether or not the value to the left of `==` is equivalent to the value to the right. For example:

```{r hw2chunk5}
x <- 4 
y <- x==4
```

-   In the next code chunk, write 2 lines of code to (1) print `y`; and (2) show the data type of `y`. [2 POINTS]

```{r hw2chunk6}

```

-   What is another word for the data type of `y`? That is, what is a word that is not returned by R but that describes the type of data held by `y`?

[ENTER YOUR ANSWER HERE, 1 POINT]

4.  Below, create a new code chunk and name it `hw2chunk7`. In the chunk, create two vectors (your choice of length). Then "column bind" the two vectors in order to create a matrix. Print the matrix and its dimensions.

[CREATE YOUR CODE CHUNK HERE, 3 POINTS]

5.  When you run the next code chunk, you will get an error.

```{r hw2chunk8}
my.object <- dog
```

-   Explain the error and make a correction in the code chunk.

[ENTER ANSWER HERE, 2 POINTS]

6.  When you run the next code chunk, you will get an error.

```{r hw2chunk9}
my.var <- (1,2,3)
```

-   Explain the error and make a correction in the code chunk.

[ENTER ANSWER HERE, 2 POINTS]

### Extra Credit

1.  Run the following code chunk.

```{r hw2chunk10}
my.vector <- c(4,3,"tree")
my.vector
```

-   What do you observe in the print of `my.vector`? Explain why you observe what you do.

[ENTER YOUR ANSWER HERE, 2 POINTS]

2.  In the code chunk below, create an object called `sunday.runs` and assign to it the first row of `run.times.by.week`. [2 POINTS]

```{r hw2chunk11}

```

\
ASSIGNMENT IS WORTH 20 POINTS (24 POINTS POSSIBLE WITH EXTRA CREDIT)
