---
title: "Class #3: Data Frames in Base R"
output:
   xaringan::moon_reader:
execute: 
  echo: true
  size: scriptsize
header-includes:
  # Adjusting output size
  - \AtBeginEnvironment{verbatim}{\scriptsize}
  
  # Adjusting code chunk size using 'listings'
  - \usepackage{listings}
  - \lstset{basicstyle=\scriptsize\ttfamily,breaklines=true}
---

## Reminder to Start Zoom Recording

## Announcements & Reminders

-   Attendance important & cameras on please
-   Six Ways to Get Help & Course Collaboration Leaders
-   Moodle & assignments debrief
-   Joing the class late? Please reach out to me.

---
title: "Introduction to Base R"
format: beamer
---

## Base R: What Beginners Need to Know

### What is Base R?

-   **Base R** refers to the core set of functions, data types, and utilities included with the default R installation.
-   It provides essential tools for working with vectors, data frames, lists, and matrices, as well as basic **data manipulation**, **statistics**, and **plotting** functions.

## Alternatives to Base R

-   **Tidyverse**: A collection of packages designed for more intuitive data manipulation and visualization. Syntax is often cleaner for larger data workflows. (**We will learn this next.**)
-   **Data.table**: A faster alternative to Base R data frames, optimized for large datasets. (**Beyond scope of class.**)
-   **Matrix-specific Packages**: For large numerical datasets, for efficient matrix operations. (**Beyond scope of class.**)

## Why Learn Base R?

-   Provides a **strong foundation**: Most R packages build upon the principles of Base R, so understanding it helps you with advanced techniques.
-   **Lightweight**: Base R functions are fast and simple for smaller or less complex tasks.
-   **Always available**: Base R is built-in, meaning you don’t need to install any additional packages.

## Some syntax and terminology (mostly a review)

-   An R **object** can store different types of data, such as:
    -   **Numbers** (e.g., `10`, `3.14`)
    -   **Text** (e.g., `"Hello"`, `"R is fun!"`)
    -   **Vectors** (e.g., a list of numbers or text)
    -   **Data Frames** (e.g., tables of data)
-   We assign values to objects using the <- operator or =. 
    - Tip: <- is very much preferred, but = also works.

```{r}
my_number <- 42
my_text = "Data Science is awesome!"
```
- R **functions** are commands that perform specific tasks in R. 
  -   A function takes *input* (called arguments) and returns *output*.
  -   You can pass objects as inputs into functions to operate on them.
  -   Some functions inspect objects by telling us info about it.
  
```{r}
# Use the 'c()' function to create a vector of numbers
numbers <- c(1, 2, 3)  
# Use the `sum()` function to sum across the vector
total <- sum(numbers)
# Inspect the object
class(numbers)
```

## Overview of Data Frames in Base R

A data frame is a two-dimensional table-like structure that can hold **columns of different types** (e.g., numeric, character, and logical).

Each column in a data frame is a vector, and the length of each vector (number of rows) must be the same across all columns.

## Built-in Data Frames in R

We use a R package in base R called `datasets` to access built-in toy data frames.

```{r}
# List all datasets from the 'datasets' package
data(package="datasets")
```

## Viewing the Data Frame

Let's inspect the data frame `mtcars` by looking at the first 6 rows using the function `head()`:

```{r mtcars_head}
head(mtcars)
```

Let's look at the structure of the data frame using `str()`:

```{r mtcars_str}
str(mtcars)
```

## Column and Row Names

```{r rowNames}
colnames(mtcars)
rownames(mtcars)
```

## Accessing Columns in Data Frames

**Like with matrices, we can use `[]` to designate a column.** But it is not a good coding practice to use numbers within the brackets. If our data frame gets rearranged, our code would be wrong. Instead, we use column names as follows:

```{r getcolumn}
mpg <- mtcars[,"mpg"] # a numeric vector
class(mpg)
mpg <- mtcars["mpg"]  # a data.frame with just one column
class(mpg)
```

## Accessing Rows in Data Frames

**We can also use brackets to get particular rows.** Note here, we need to put the row name before a comma. And it returns a data.frame.

```{r getrow}
ferrari.dino <- mtcars["Ferrari Dino",] # a data.frame with just one row
class(ferrari.dino)
```

## Subsetting Rows

We can also get multiple rows:

```{r getmultrows}
# First I will create a vector that includes all Mazdas
Mazdas <- c("Mazda RX4","Mazda RX4 Wag")

# Then I will subset the data frame to the Mazdas 
mtcars.Mazdas <- mtcars[Mazdas,]

# Print
print(mtcars.Mazdas)
```

## More Advanced Row Subsetting

Let's get all the Mercedes without having to type them out.

```{r getmultrows_substr}
# Get the first four characters of the row names 
sub4 <- substr(rownames(mtcars),1,4)
# Create a logical vector that is equal to TRUE when sub4 is equal to "Merc"
is.Merc <- sub4=="Merc"
# Check
is.Merc
# Make a new data frame that is just the rows for Mercedes cars
mtcars.Mercedes <- mtcars[is.Merc,]
# Print
mtcars.Mercedes
```

## Using \$ to Access Columns

**We can also use "data.frame.name\$" to refer to columns/variables.**

```{r showdollarsign}
# The object mpg is being assigned the column names "mpg"
mpg <- mtcars$mpg
# Let's look at the mean value of mpg
mean(mpg)
```

## Creating a New Column

We can also use this approach to add columns to a data frame. In the example below, we create a new column `kpg` that converts miles per gallon (`mpg`) to kilometers per gallon.

```{r createcol}
# note since we are adding something new, rather than changing or deleting, it is fine to keep the same object name for this change
mtcars$kpg <- mtcars$mpg * 1.6
head(mtcars)
```

## Reading and Writing Data Files

### .RData (or .Rda) files

-   .RData files can store multiple R objects. Sometimes data files are stored as .rds files, which only store a single object.

-   After loading, the objects contained in the .RData file will be available in your R environment with their original names.

```{r loadSleepStudy}
load("data/SleepStudy.Rda")  # Load a .RData file
# save(my.object, file = "path/output.Rda")  # Save as .RData
```

### .csv files or .txt files

A common way to import external data into R is from a CSV file. Use the read.csv() function.

```{r showcsv}
#data <- read.csv("path/file.csv")
# write.csv(my.data, "path/output.csv")  # Write to a .csv file

```

For plain text files, use read.table() or readLines():

```{r showtxt}
# data <- read.table("path/file.txt", header = TRUE)
# lines <- readLines("path/file.txt")
# write.table(my.data, "path/output.txt", row.names = FALSE)
```

## Tips

-   File Paths: File paths can be absolute or relative.
    -   Absolute paths specify the exact location on your computer.
    -   Relative paths are relative to your current working directory in R.
    -   **We set up R Studio so that it should be easy to use relative paths!**

-   Working Directory: 
    -   Use `getwd()` to find your current working directory
    -   Use `setwd("path/to/directory")` to change it.
    -   **But we set up R Studio so that if you `.Rproj` is loaded, your current working directory is set.**

-   Viewing Data: After reading a file, use head(data) to view the first few lines of your data frame.

-   Handling Errors: If R can't read your file, check for typos in the file path and ensure the file format is correct.
