---
title: "Introduction to Data Visualization in R"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ggplot2)
```

## Introduction to Data Visualization in R

In this notebook, we will explore how to create basic visualizations using R. This notebook builds on the lessons you have learned in using the dplyr. We will focus on plotting capabilities in base R as well as in and the `ggplot2` library, which is part of the tidyverse.

### Python vs. R for data visualization

Both Python and R offer robust data visualization capabilities, with each language having its own strengths and challenges. Python's ecosystem is diverse and well-suited for integration with web-based applications, while R's ggplot2 package offers a powerful and expressive syntax for creating publication-quality graphics. Ultimately, the choice between Python and R for data visualization depends on factors such as personal preference, project requirements, and existing expertise in each language.

### Examples of Basic Visualizations

In this notebook, we will explore bar charts, histograms, scatter plots, and line plots. We will look at how to do this with base R as well as with `ggplot2`. 

### Loading the dataset

Let's load the colleges dataset - the same one we looked at for Python data viz. 

```{r load-data}
collegesDat <- read_csv('data/colleges.csv')
```

## (1) Bar Charts

#### Basic R

In base R, we use the `barplot()` function. The help documentation provides lots of assistance with different specifications to make your plot more readable. 

```{r bar-chart-basic}
highest_degree_counts <- table(collegesDat$highest_degree)
barplot(highest_degree_counts,
        main="Distribution of Colleges by Highest Degree Offered",
        xlab="Highest Degree Offered",
        ylab="Number of Colleges",
        col="blue")
```

#### ggplot2

Here is how we make the same plot using `ggplot2`. 

```{r bar-chart-ggplot}
ggplot(collegesDat, aes(x=highest_degree)) +
  geom_bar(fill="blue") + # this specifies to create a bar chart
  theme_minimal() +  # this specifies a set of formatting options 
  labs(title="Distribution of Colleges by Highest Degree Offered", x="Highest Degree Offered", y="Count")
```

## (2) Histograms

#### Basic R

The hist function in R creates a histogram. Again see `help(hist)` for info on specifications for tailoring your plot. 

```{r histogram-basic}
hist(collegesDat$grad_rate, main="Histogram of Graduation Rates", xlab="Graduation Rate", breaks=10)
```

#### ggplot2

In ggplot2, we start the same way we did for the bar plot (the first line is exactly the same), but now we use `geom_histogram()`. 

```{r histogram-ggplot}
ggplot(collegesDat, aes(x=grad_rate)) +
  geom_histogram(binwidth=10,fill="dark green") +
  labs(title="Histogram of Graduation Rates", x="Graduation Rate", y="Frequency")
```


## (3) Scatter Plots

#### Basic R

For a quick plot in base R, simply use the `plot()` function. 

```{r scatter-basic}
plot(collegesDat$avg_cost, collegesDat$grad_rate, main="Graduation Rates vs. Average Cost", xlab="Average Cost", ylab="Graduation Rate", col="blue")
```

#### ggplot2

But here, let's apply the grammar of graphics so that we can see differences by ownership of the colleges. We will add this as a third aesthetic, assigning college ownership to color. Here, I am going to assign my plot to an object. This will allow me to add to it later, without having to repeat everything. When I do this, I need to include a print command. 

```{r scatter1-ggplot}
scatter1 <- ggplot(collegesDat, aes(x=avg_cost, y=grad_rate, color=ownership)) +
  geom_point() +
  labs(title="How Graduation Rates Vary by Average Cost", x="Average Cost", y="Graduation Rate")
print(scatter1)
```


So let's try to make it prettier. Here are some steps towards that. Note, you can find a set of color options here[https://r-graph-gallery.com/ggplot2-color.html].

```{r scatter2-ggplot}
scatter2 <- scatter1 +
            theme_minimal() + # using theme_minimal to change look of plot
            theme(legend.position = c(0.8, 0.8)) + # moves the legend to the upper right
            scale_color_manual(values = c("Private nonprofit" = "darkorchid4", "Prviate for-profit" = "darkseagreen3", "Public" = "darksalmon")) 
print(scatter2)
```
Here I want to add a new variation of the scatter plot - a bubble plot. This simply uses the SIZE of the points to add a third variable's value to the plot. For example, we could use the size to indicate the enrollment. Here's an example: 


```{r bubbleplot}
# Let's restrict our analysis just to private schools in which the highest degree is a Bachelors degree
filterDat <- collegesDat %>%
  filter(state == "NC") 

# Bubble plot
ggplot(filterDat, aes(x = admit_rate, y = grad_rate, size = enrollment)) +
  geom_point(alpha = 0.6) +
  scale_size_continuous(range = c(1, 20)) +
  labs(title = "Graduation Rate vs. Admission Rate by Enrollment Size for all Colleges in North Carolina",
       x = "Admission",
       y = "Graduation Rate",
       size = "Enrollment") +
  theme_minimal() +
  theme(legend.position = "right") + 
  scale_size_continuous(range = c(1, 20),
                        breaks = c(500, 1000, 5000, 10000, 15000, 20000), # Adjust breaks to suitable enrollment values
                        labels = c("500", "1k", "5k", "10k", "15k", "20k"))  # Label adjustments for readability
  
```

## (4) Line Plots

Here we will use the UFO data. Like we did in python, we need to first get counts of sightings by hear. We can do this with `dplyr`. 

```{r loadUFOdata}
UFOdat <- read_csv("data/ufo_sightings.csv")

sightings_by_year <- UFOdat %>%
  group_by(Dates.Documented.Year) %>%
  summarise(Count = n())

# To view the result
print(sightings_by_year)
```
Next, let's plot with `ggplot2`. 

```{r line-plot2}
ggplot(sightings_by_year, aes(x=Dates.Documented.Year, y=Count)) +
  geom_line(color="purple") +
  theme_classic() + 
  labs(title="UFO Sightings by Year", x="", y="")
```

Finally, let's replicate what we did in Python to show a line plot for each of the states, highlight just one state to compare with the rest. 

```{r lineplot2}
# First, we need sightings by year and state
sightings_by_year_state <- UFOdat %>%
  group_by(Dates.Documented.Year, Location.State) %>%
  summarise(Count = n())

# Let's check our output
head(sightings_by_year_state)

# Let's pick a state to highlight
highlight_state <- 'CA'

# And plot note, we use the aes of group
ggplot(sightings_by_year_state, aes(x = Dates.Documented.Year, y = Count, group = Location.State)) +
  geom_line(aes(color = Location.State == highlight_state), alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "UFO Sightings by Year, Highlighting California",
       x = "Year",
       y = "Number of Sightings") +
  theme_minimal() +
  theme(legend.position = "none") # Hide the legend


```

## (5) Heat Maps

A heatmap is a graphical representation of data in which values in a matrix are represented as shades of or different colors. They are designed so that viewers grasp patterns, variances, and relationships in the data quickly through color gradients.

Heat maps can be used to reveal...

Patterns: Easily identify trends and patterns over time or across categories. For example, in a heatmap of website traffic by day and hour, you might see higher traffic (darker colors) during weekdays and working hours.
Anomalies: Spot outliers or anomalies that deviate from the overall pattern. In a heatmap of sensor data across different locations, an unusually high temperature at a single sensor location would stand out.
Correlations: Understand the correlation between different variables. A heatmap of correlation coefficients between financial assets could highlight which assets move together or in opposition.
Density: Visualize the density or intensity of events. A heatmap of crime incidents in a city could show areas with higher crime rates.
Comparisons: Compare different categories or groups. In sales data across different regions and products, a heatmap can quickly show which products are bestsellers in which regions.


Here are some example of when heatmaps can be valuable: 

Weather Data Visualization: Heatmaps are used to visualize temperature changes across different geographical regions, with colors representing temperature ranges. This helps in understanding climate patterns or the impact of a heatwave.

Genomic Data Analysis: In bioinformatics, heatmaps are used to display gene expression data, where rows represent genes and columns represent different experimental conditions. The color intensity indicates the level of gene expression, helping in identifying genes that are overexpressed or underexpressed under certain conditions.

Financial Market Data: Heatmaps are employed to show the performance of stocks in a market, with each cell representing a stock. The color shades may indicate the stock's daily or yearly performance, helping investors identify sectors or stocks that are performing well or poorly.

Website Traffic Analysis: Analyze web page activity by showing the frequency of visits, clicks, or other actions on different parts of a page or across various times of the day. This helps in optimizing website design and content placement for better user engagement.

Educational Data: In the context of colleges, a heatmap could be used to visualize variables such as graduation rates, median alumni earnings, or student satisfaction across different departments or programs. This can highlight areas of strength and opportunities for improvement within an institution.

Let's make a simple example with out colleges data. Let's examine average median debt by region and by ownership of the colleges. First, we need to get the average of median debt within each combination of region and ownership: 
```{r heatmapPrep}
# First I'm renaming one region for easier printing
collegesDat$region[collegesDat$region=="Rockies & Southwest"] <- "Southwest"

# Here we filter and group data -then summarize to get the average
heatmap_data <- collegesDat %>%
  filter(region != "US Service Schools") %>%
  group_by(region,ownership) %>%
  summarise(avg_median_debt = mean(median_debt, na.rm = TRUE)) %>%
  ungroup()

# View the first few rows of the aggregated data
head(heatmap_data)
```

Then we can use ggplot2 for creating our heatmap. Notice we use "geom_tile" for the heatmap. Also notice that we use region and ownership for the x and y axes and the average median debt for "fill."

```{r heatmap}
ggplot(heatmap_data, aes(x = region, y = ownership, fill = avg_median_debt)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Average Median Debt") +
  labs(title = "Average Median Debt by Region and Ownership", x = "Region", y = "Ownership") +
  theme_minimal() 
```

As mentioned above, heat maps can be helpful to look at correlations across a group of variables. Let's give this a try: 

```{r corr}
# This will select only the columns that are of type double (continuous)
double_vars <- select_if(collegesDat, is.double) # using dplyr

# Compute correlations - this returns a matrix with each row a variable and each column a variable
corCollegeVars <- cor(double_vars, use = "complete.obs")

# We need to reshape
library(reshape2)
melted_corr_matrix <- melt(corCollegeVars)

# Plot with ggplot2
library(ggplot2)

# Create the heatmap
ggplot(data = melted_corr_matrix, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile() +  # This creates the tiles for the heatmap
  scale_fill_gradient2(low = "green", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +  # Optional: Use a minimal theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x axis labels
  labs(x = "", y = "", title = "Correlation Matrix Heatmap")  # Customize labels and title




```