---
title: "Introduction to Data Visualization in R"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ggplot2)
```

## Introduction to Data Visualization in R

In this notebook, we will explore how to create basic visualizations using R. This notebook builds on the lessons you have learned in using the dplyr. We will focus on plotting capabilities in base R as well as in and the `ggplot2` library, which is part of the tidyverse.

### Python vs. R for data visualization

Both Python and R offer robust data visualization capabilities, with each language having its own strengths and challenges. Python's ecosystem is diverse and well-suited for integration with web-based applications, while R's ggplot2 package offers a powerful and expressive syntax for creating publication-quality graphics. Ultimately, the choice between Python and R for data visualization depends on factors such as personal preference, project requirements, and existing expertise in each language.

### Examples of Basic Visualizations

In this notebook, we will explore bar charts, histograms, scatter plots, and line plots. We will look at how to do this with base R as well as with `ggplot2`. 

### Loading the dataset

Let's load the colleges dataset - the same one we looked at for Python data viz. 

```{r load-data}
collegesDat <- read_csv('data/colleges.csv')
```

## (1) Bar Charts

#### Basic R

In base R, we use the `barplot()` function. The help documentation provides lots of assistance with different specifications to make your plot more readable. 

```{r bar-chart-basic}
highest_degree_counts <- table(collegesDat$highest_degree)
barplot(highest_degree_counts,
        main="Distribution of Colleges by Highest Degree Offered",
        xlab="Highest Degree Offered",
        ylab="Number of Colleges",
        col="blue")
```

#### ggplot2

Here is how we make the same plot using `ggplot2`. 

```{r bar-chart-ggplot}
ggplot(collegesDat, aes(x=highest_degree)) +
  geom_bar(fill="blue") + # this specifies to create a bar chart
  theme_classic() +  # this specifies a set of formatting options 
  labs(title="Distribution of Colleges by Highest Degree Offered", x="Highest Degree Offered", y="Count")
```

## (2) Histograms

#### Basic R

The hist function in R creates a histogram. Again see `help(hist)` for info on specifications for tailoring your plot. 

```{r histogram-basic}
hist(collegesDat$grad_rate, main="Histogram of Graduation Rates", xlab="Graduation Rate", breaks=10)
```

#### ggplot2

In ggplot2, we start the same way we did for the bar plot (the first line is exactly the same), but now we use `geom_histogram()`. 

```{r histogram-ggplot}
ggplot(collegesDat, aes(x=grad_rate)) +
  geom_histogram(binwidth=10,fill="dark green") +
  labs(title="Histogram of Graduation Rates", x="Graduation Rate", y="Frequency")
```


## (3) Scatter Plots

#### Basic R

For a quick plot in base R, simply use the `plot()` function. 

```{r scatter-basic}
plot(collegesDat$avg_cost, collegesDat$grad_rate, main="Graduation Rates vs. Average Cost", xlab="Average Cost", ylab="Graduation Rate", col="blue")
```

#### ggplot2

But here, let's apply the grammar of graphics so that we can see differences by ownership of the colleges. We will add this as a third aesthetic, assigning college ownership to color. Here, I am going to assign my plot to an object. This will allow me to add to it later, without having to repeat everything. When I do this, I need to include a print command. 

```{r scatter1-ggplot}
scatter1 <- ggplot(collegesDat, aes(x=avg_cost, y=grad_rate, color=ownership)) +
  geom_point() +
  labs(title="How Graduation Rates Vary by Average Cost", x="Average Cost", y="Graduation Rate")
print(scatter1)
```


So let's try to make it prettier. Here are some steps towards that. Note, you can find a set of color options here[https://r-graph-gallery.com/ggplot2-color.html].

```{r scatter2-ggplot}
scatter2 <- scatter1 +
            theme_minimal() + # using theme_minimal to change look of plot
            theme(legend.position = c(0.8, 0.8)) + # moves the legend to the upper right
            scale_color_manual(values = c("Private nonprofit" = "darkorchid4", "Prviate for-profit" = "darkseagreen3", "Public" = "darksalmon")) 
print(scatter2)
```
Here I want to add a new variation of the scatter plot - a bubble plot. This simply uses the SIZE of the points to add a third variable's value to the plot. For example, we could use the size to indicate the enrollment. Here's an example: 


```{r bubbleplot}
# Let's restrict our analysis just to private schools in which the highest degree is a Bachelors degree
filterDat <- collegesDat %>%
  filter(state == "NC") 

# Bubble plot
ggplot(filterDat, aes(x = admit_rate, y = grad_rate, size = enrollment)) +
  geom_point(alpha = 0.6) +
  scale_size_continuous(range = c(1, 20)) +
  labs(title = "Graduation Rate vs. Admission Rate by Enrollment Size for all Colleges in North Carolina",
       x = "Admission",
       y = "Graduation Rate",
       size = "Enrollment") +
  theme_minimal() +
  theme(legend.position = "right") + 
  scale_size_continuous(range = c(1, 20),
                        breaks = c(500, 1000, 5000, 10000, 15000, 20000), # Adjust breaks to suitable enrollment values
                        labels = c("500", "1k", "5k", "10k", "15k", "20k"))  # Label adjustments for readability
  
```

## (4) Line Plots

Here we will use the UFO data. Like we did in python, we need to first get counts of sightings by hear. We can do this with `dplyr`. 

```{r loadUFOdata}
UFOdat <- read_csv("data/ufo_sightings.csv")

sightings_by_year <- UFOdat %>%
  group_by(Dates.Documented.Year) %>%
  summarise(Count = n())

# To view the result
print(sightings_by_year)
```
Next, let's plot with `ggplot2`. 

```{r line-plot2}
ggplot(sightings_by_year, aes(x=Dates.Documented.Year, y=Count)) +
  geom_line(color="purple") +
  theme_classic() + 
  labs(title="UFO Sightings by Year", x="", y="")
```

Finally, let's replicate what we did in Python to show a line plot for each of the states, highlight just one state to compare with the rest. 

```{r lineplot2}
# First, we need sightings by year and state
sightings_by_year_state <- UFOdat %>%
  group_by(Dates.Documented.Year, Location.State) %>%
  summarise(Count = n())

# Let's check our output
head(sightings_by_year_state)

# Let's pick a state to highlight
highlight_state <- 'CA'

# And plot note, we use the aes of group
ggplot(sightings_by_year_state, aes(x = Dates.Documented.Year, y = Count, group = Location.State)) +
  geom_line(aes(color = Location.State == highlight_state), alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "UFO Sightings by Year, Highlighting California",
       x = "Year",
       y = "Number of Sightings") +
  theme_minimal() +
  theme(legend.position = "none") # Hide the legend


```
