---
title: "R Notebook #2: Dive into Data Frames"
output: pdf_document
---

# Overview of data frames

A data frame is a two-dimensional table-like structure that can hold **columns of different types** (e.g., numeric, character, and logical). Each column in a data frame can be thought of as a vector, and the length of each vector (number of rows) must be the same across all columns.

```{r dataframes}
# Here, first I am loading an R "package" called "datasets." R packages are the building blocks on top of "base R." They are contributed by wide-ranging organizations and individuals. The datasets package includes many "built-in" datasets that you can use to explore R. When you load the package with "libary(datasets)" all of the datasets in the package become available. 
library(datasets)

# We can look at the structure mtcars with str() (Note this function works for other types of objects as well. The info returned depends on the type of object.)
str(mtcars)
```

Let's look at the first 6 rows of the data frame mtcars, by using the function `head()`

```{r mtcars_view}
head(mtcars)
```

You can see that the car types do not have a column name. This is because the car names are not actually a column in the data frame. Instead, the rows also have names, which are the car types.

```{r rowNames}
# here we can check the column names to see the variables in the data frame
colnames(mtcars)

# here we can see the row names
rownames(mtcars)
```

# Navigating R data frames with base R

You can access a data frame's elements in several ways:

(1) Like with matrices, we can use `[]` to designate a column. But it is not a good coding practice to use numbers within the brackets. If our data frame gets rearranged, our code would be wrong. But we can use column names as follows:

```{r getcolumn}
mpg <- mtcars["mpg"]  # here mpg is a data.frame with just one column
mpg <- mtcars[,"mpg"] # here mpg is a numeric vector
```

We can also use brackets to get particular rows. Note here, we need to put the row name before a comma. And it returns a data.frame.  
```{r getrow}
ferrari.dino <- mtcars["Ferrari Dino",] # a data.frame with just one row
```

Here's a little expansion in which I get all the Mercedes. 

```{r getmultrows}
# First I will assign the rownames to a character vector:
car.types <- rownames(mtcars)
```


Accessing the first two rows: R Copy code first_two_rows \<- my_data[1:2, ] Accessing Specific Elements: Combine row and column indices.

Accessing a specific element (row 2, column Name): R Copy code second_name \<- my_data[2, "Name"]
